Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-05-27 submission/Sze_Obesity_PeerJ_2016.Rmd

**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 10 studies that fit our criteria. We applied three analysis pipelines to the data from each study. First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value. We also used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level. Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F. Finally, we utilized linear mixed-effect models to correct for variation by study and analyzed the pooled community diversity and B/F values within each dataset to test for an association with obesity.

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese individuals. A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)). With individual studies, random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05). When we pooled the results across studies, there was a significant difference in RR of obesity based on Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`). When we used a linear mixed-effect model correcting for study there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals. Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome and its effect on obesity. Our meta-analysis provides support for the observation that there is a difference between non-obese and obese individuals in Shannon diversity but that the effect sizes are small, and that a very large number of indivdiuals would have to be tested to reach sufficient study power.

@@

***Literature Review and Study Inclusion***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement [Figure 1]. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified ten additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The ten studies are summarized in Tables 1 and 2.

@@

***Data Analysis***
We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). We split the overall analysis into three general strategies. In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, observed richness, and evenness. A two-tailed Wilcoxon test was performed for comparison for non-obese and obese individuals. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan R package. Finally, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.  To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.  A Bonferroni-adjusted p-value was used to assess differences between each study.

@@

In the third strategy, we corrected for study effects and used linear mixed-effect models to test the pooled diversity and B/F ratio data for a difference between obese and non-obese groups. Additionally, we also performed power and sample number simulations for different effect sizes for each study.  This was completed for all the different components measured as part of first strategy.  We also calculated the actual sample size needed based on the effect size of each individual study.  To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.

@@

***Separate Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity, three for observed richness, one for evenness, and one for Bacteroidetes [Table S1]. There were no significant results for Firmicutes or the B/F ratio. The full summary of the exact P-values for each respective study and variable tested can be found in table S1 of the online supplement.

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm. We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.  We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al. study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al. Study).  For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al. study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al. study). The total number of trees and node size was standardized to 500 and 20 respectively for each data set. Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05, Figure 2).

@@

We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis. From the graph we can observe that there does not seem to be any significant  bias [Figure S1]. Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.

***Linear Mixed-Effect models for Pooled Analysis***

Using the third approach which utilizes a linear mixed-effect model to control for the effect of study on the Shannon diversity  we found that there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 5A]. Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`).  However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) using this approach, between obese and non-obese individuals [Figure 5B].

@@

For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 5A] compared to the B/F ratio [Figure 6A] differences.  When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 5A].  This predicted power was also similar for evenness (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`).   However, this was not the case for the B/F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 6A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.   Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 5B].  In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 6B].  A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S2-S7].

>>>>

**Methods:** To identify datasets that we included in the meta analysis we followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines we winnowed an initial set of 196 studies that tested the microbiome-obesity association hypothesis down to 10 studies that fit our criteria. We applied three analysis pipelines to the data from each study. First, we tested the association between community structure and diversity and the Bacteroides to Firmicutes ratio (B/F) with the subjects' body mass index (BMI) treated as a continuous or categorical value. We investigated each study seperately as well as together while correcting for study effects.  Second, we pooled the relative risk (RR) ratio of being obese based on characteristics of the microbiome for each study and tested for an association between RR and community diversity and B/F. Finally, We also used the random forest machine learning algorithm to identify community features that would allow us to to classify whether a person was obese based on either genus or OTU level and assessed how well these variables did across study.  Additionally, we also investigated power and esimated sample sizes needed to adequately test our observer effect sizes amongst the different study populations.

**Results:** When we considered each dataset in isolation, there were no significant differences (P > 0.05) in B/F between non-obese and obese groups. A total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric != "bacteroidetes"), ]))` diversity measurements  were found to be significantly different between non-obese and obese indivdiuals (P < 0.05) (Shannon Diversity (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannon"), ]))`), observed richness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "sobs"), ]))`), and evenness (n=`r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05 & alpha_tests.summary$metric == "shannoneven"), ]))`)).  When we used a linear mixed-effect model correcting for study effects there was a significant difference in Shannon diversity (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 8), nsmall=8)`) but no such significant difference was found for the B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`) between non-obese and obese individuals.  When we pooled the results across studies and looked at whether there was an increased RR of obesity we found that there was for Shannon diversity (RR = `r format(round(RR_composite['shannon', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['shannon', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['shannon', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['shannon', 'p_value'], 3), nsmall=3)`) but not for B/F ratio (RR = `r format(round(RR_composite['bf_ratio', 'rr'],2), nsmall=2)`, CI = `r format(round(RR_composite['bf_ratio', 'ci_lb'], 2), nsmall=2)`-`r format(round(RR_composite['bf_ratio', 'ci_ub'], 2), nsmall=2)`, P-value = `r format(round(RR_composite['bf_ratio', 'p_value'], 3), nsmall=3)`).  Overall random forest was effective at identifying features within the datasets for differentiating between non-obese and obese individuals for both the genus (mean AUC: `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)`) and OTU (mean AUC: `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)`; s.d.=`r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)`) level, and there was no significant difference between the studies (bonferroni adjusted P-value > 0.05).  Finally, when completing both power and sample size calculations and simulations we found that most of the studies examined were not sufficiently powered to detect differences in any of the measurements used.

**Conclusion:** We show that by pooling the results of many individual studies we can test the generalizability of the microbiome and its effect on obesity. Our meta-analysis provides support for the observation that there is a difference between non-obese and obese individuals in Shannon diversity but that the effect sizes is small, and that a very large number of indivdiuals would have to be tested to reach a sufficient study power.

@@

***Literature Review and Study Inclusion***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified ten additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The ten studies are summarized in Tables 1 and 2.

@@

***Data Analysis***
We performed the analysis using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). We split the overall analysis into three general strategies. In the first strategy we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B/F), Shannon diversity, observed richness, and evenness. After each variable was transformed to fit a normal distribution A two-tailed t-test was performed for comparison for non-obese and obese individuals. Additionaly, for this strategy, we performed a pooled analysis on these measured variables using linear mixed effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups.  Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan R package.

@@

In the third strategy, we used the AUCRF (v1.1) R package to implement a Random Forest analysis to identify the best OTUs and alpha-diversity metrics for distinguishing non-obese and obese individuals.  To generate Receiver Operator Characteristic (ROC) curves we used the pROC (v1.8) R package.  A Bonferroni-adjusted p-value was used to assess differences between each study.

Additionally, we also performed power and sample number simulations for different effect sizes for each study.  This was completed for all the different components measured as part of first strategy.  We also calculated the actual sample size needed based on the effect size of each individual study.  To complete this analysis we used a combination of the pwr (v1.1-3) R package and base R functions.

@@

***Separate and Pooled Analysis***
We used the categorical variable of obese or non-obese to analyze each data set separately for significant correlations to Bacteroidetes, Firmicutes, Bacteroidetes/Firmicutes (B/F) ratio, Shannon diversity, OTU richness, evenness, and the Bray-Curtis distance matrix which was a similar approach taken by two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Using a P-value cutoff for significance as less than 0.05 we found a total of `r length(rownames(alpha_tests.summary[which(alpha_tests.summary$p.value < 0.05), ]))` significant results; two for Shannon diversity [Figure 2A], three for observed richness [Figure S1], one for evenness [Figure S2], and one for Bacteroidetes [Figure S3]. There were no significant results for Firmicutes [Figure S4] or the B/F ratio [Figure 2B]. Using the linear mixed-effect model to analyze the data pooled together while also controling study effects we found that for Shannon diversity  there was a significant difference between the obese and non-obese groups (P-value = `r format(round(alpha_composite.summary['shannon', 'composite_p'], 4), nsmall=4)`) [Figure 2A]. Additionally,  there was also a significant difference in obseverd richness (P-value = `r format(round(alpha_composite.summary['sobs', 'composite_p'], 9), nsmall=9)`) [Figure S1].  However, we found that the there was no significant difference in evenness (P-value = `r format(round(alpha_composite.summary['shannoneven', 'composite_p'], 2), nsmall=2)`), Bacteroidetes (P-value = `r format(round(alpha_composite.summary['bacteroidetes', 'composite_p'], 2), nsmall=2)`), Firmicutes (P-value = `r format(round(alpha_composite.summary['firmicutes', 'composite_p'], 2), nsmall=2)`), and B/F ratio (P-value = `r format(round(alpha_composite.summary['bf_ratio', 'composite_p'], 2), nsmall=2)`)  [Figure 2B & S2-S4] using this approach, between obese and non-obese individuals.

@@

We also used a funnel plot to asses the bias of the 9 data sets used in this meta-analysis. From the graph we can observe that there does not seem to be any significant  bias [Figure S5]. Both funnel plots have almost equal number of studies scattered on either side of the estimated true value.

***Random Forest Classification and Predicition Accuracy***

Next we investigated how well the OTUs in each study were able to classify obese and non-obese individuals using the random forest algorithm. We found that the average Out Of Bag (OOB) Area Under the Curve (AUC) was `r format(round(mean(rand_forest.genus.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.genus.summary$auc), 2), nsmall=2)` for the genus level analysis and `r format(round(mean(rand_forest.otu.summary$auc), 2), nsmall=2)` +/- `r format(round(sd(rand_forest.otu.summary$auc), 2), nsmall=2)` for the OTU level analysis.  We found that the range for the total number of variables used for the classification using genus ranged from `r min(rand_forest.genus.summary$k_opt)` (HMP and Wu, et al. study) to `r max(rand_forest.genus.summary$k_opt)` (Goodrich, et al. Study).  For the OTU level the number of variables ranged from `r min(rand_forest.otu.summary$k_opt)` (Wu, et al. study) to `r max(rand_forest.otu.summary$k_opt)` (Goodrich, et al. study). The total number of trees and node size was standardized to 500 and 20 respectively for each data set. Overall there was no significant difference between the ROC curves of each data set (P-value > 0.05) regardless if it was using genus or OTU level classification [Figure 4].  Overall most studies performed poorly when used on other data sets for classification [Figure 5] with no significant difference between studies for their mean accuracy for prediction of non-obese and obese indivdiuals (P-value > 0.05).

@@

For Shannon Diversity the measured difference between both obese and non-obese groups was relatively small [Figure 6A] compared to the B/F ratio [Figure 7A] differences.  When simulating different effect size differences of 1%, 5%, 10%, and 15% we found that many studies would be powered sufficiently to find a 15% difference in Shannon diversity (predicted percent power = `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`) [Figure 6A].  This predicted power was also similar for evenness (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`, SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannoneven' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`).   However, this was not the case for the B/F ratio (`r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)` +/- `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.15), ]$power*100), 2), nsmall=2)`%) [Figure 7A] or the other measures that were tested (observed richness, Bacteroidetes, Firmicutes) where even a 15% difference between non-obese and obese groups only resulted in a relatively poor average power between studies.   Looking at the sample sizes showed that on average to detect a 5% difference in Shannon diversity at 80% power between non-obese and obese individuals a total n of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'shannon' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) would be needed for each group [Figure 6B].  In contrast in order to detect a 5% difference with 80% power in the B/F ratio a sample size of `r format(round(mean(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)` (SD = `r format(round(sd(alpha_power.predicted[which(alpha_power.predicted$metric == 'bf_ratio' & alpha_power.predicted$effect_size == 0.05), ]$balanced_n*100), 0), nsmall=0)`) per group [Figure 7B].  A complete breakdown of the power simulations and sample size estimates for the other variables investigated and RR can be found in the online supplement [Figure S6-S11].
